---
title: "Crude Oil Price Predictor 2020"
author: "Christopher Page"
date: "8/12/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction/Overview/Executive Summary

Oil is an important commodity.  Some of our planet's inhabitants produce it.  Many more consume it. Because of its wide-ranging effects, oil commands collective attention enabled by continuous analysis.

One key metric to take into account in the continuous analysis of the oil indutry is the daily closing price of crude oil futures traded on a global basis in such venues as the New York Mercantile Exchange. 

Methods for predicting that price may entail studying the quantitative and qualitative factors at play inside the oil industry as well as in such adjacent industries as shipping and manufacturing.

They may also entail studying the complex geopolitical, military, social, and environmental developments that help explain why crude oil closing prices rise and fall as they do over time. 

This project proposes a simpler method, one that predicts crude oil closing prices based on the closing prices of complementary (e.g., gasoline) and competing (e.g., platinum) commodities.

The premises is that there is much to be learned by studying the behaviors of commodity traders engaged in making investment decisions based on running evaluations of risks and rewards. 

```{r cars}
summary(cars)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
