---
title: "Crude Oil Price Predictor 2020"
author: "Christopher Page"
date: "8/12/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction/Overview/Executive Summary

Oil is an important commodity.  Some of our planet's inhabitants produce it.  Many more consume it. All are affected by it.  Because of its wide-ranging effects, oil commands collective attention enabled by continuous analysis.

One of the key metrics to take into account in the continuous analysis of the oil indutry is the daily closing price of crude oil futures traded on a global basis in such venues as the New York Mercantile Exchange. 

Methods for predicting that price may entail studying the quantitative and qualitative factors at play inside the oil industry as well as in such adjacent industries as transportation and manufacturing.

They may also entail studying the often complex diplomatic, military, economic, cultural, and environmental developments that help explain why crude oil closing prices rise and fall as they do over time. 

This project proposes a simpler method, one that predicts crude oil closing prices based on time and the closing prices of complementary (e.g., gasoline) and competing (e.g., platinum) commodities.

The premise is that there is much to be learned by studying the behaviors of commodity traders engaged in making daily investment decisions based on running evaluations of risks and rewards. 

That premise informed the development of algorithms with the potential to predict crude oil closing prices with a viable level of accuracy, quantified as the Root Means Square Error (RMSE).

Using a 25,190-row data set constructed with publicly available information downloaded for free from https://www.nasdaq.com/, the author developed and then evaluated fifteen such algorithms.

Evaluation led to the selection of one algorithm that generated the lowest RMSE when applied to a test set. That algorithm used a Ranger model which took into account six key predictors:

(1) Year
(2) Month 
(3) Closing Price of Heating Oil
(4) Closing Price of Gasoline
(5) Closing Price of Platinum
(6) Closing Price of Soybeans

The RMSE in question was 2.92, a figure equating to 12% of the 23.51 generated by an algorithm that only took into account the average closing price across the period of observation.

## Method/Analysis

```{r load libraries, echo = FALSE}

### This part of the script loads the necessary libraries from the repository http://cran.us.r-project.org

if(!require(caret)) install.packages("caret", repos = "http://cran.us.r-project.org")
if(!require(caTools)) install.packages("caTools", repos = "http://cran.us.r-project.org")
if(!require(dplyr)) install.packages("dplyr", repos = "http://cran.us.r-project.org")
if(!require(dslabs)) install.packages("dslabs", repos = "http://cran.us.r-project.org")
if(!require(forcats)) install.packages("forcats", repos = "http://cran.us.r-project.org")
if(!require(foreach)) install.packages("foreach", repos = "http://cran.us.r-project.org")
if(!require(gam)) install.packages("gam", repos = "http://cran.us.r-project.org")
if(!require(ggplot2)) install.packages("ggplot2", repos = "http://cran.us.r-project.org")
if(!require(ggrepel)) install.packages("ggrepel", repos = "http://cran.us.r-project.org")
if(!require(ggthemes)) install.packages("ggthemes", repos = "http://cran.us.r-project.org")
if(!require(lubridate)) install.packages("lubridate", repos = "http://cran.us.r-project.org")
if(!require(purrr)) install.packages("purrr", repos = "http://cran.us.r-project.org")
if(!require(randomForest)) install.packages("randomForest", repos = "http://cran.us.r-project.org")
if(!require(tidyverse)) install.packages("tidyverse", repos = "http://cran.us.r-project.org")

```

```{r load download and wrangle data, echo = FALSE}

### This part of the script reads "data/commodity_closing_prices_2010_2020.csv" into a new data frame named "commodities"
### Then, it mutates the resulting data set in way that will enable the envisioned analysis
### Next, it saves that data frame into a .RDATA data set under the RDA subdirectory

commodities <- read_csv("data/commodity_closing_prices_2010_2020.csv")

commodities <- commodities %>%
  rename(date = Date) %>%
  mutate(date = as.Date(mdy(date))) %>%
  mutate(date_year = year(date)) %>%
  mutate(date_quarter_of_the_year = quarter(date)) %>%
  mutate(date_month = round_date(date, unit = "months")) %>%
  mutate(date_month_of_the_year = month(date)) %>%
  mutate(date_day = day(date)) %>%
  mutate(date_weekday = weekdays(date)) %>%
  rename(commodity = Commodity) %>%
  rename(sector = Sector) %>%
  rename(closing_price = `Closing Price`)

save(commodities, file = "rda/commodities.rdata")

### This part generates and saves data sets focused on the energy, precious metals, and agricultural sectors

energy <- commodities %>%
  filter(sector == 'Energy')
save(energy, file = "rda/energy.rdata")

precious_metals <- commodities %>%
  filter(sector == 'Precious Metals')
save(precious_metals, file = "rda/precious_metals.rdata")

agriculture <- commodities %>%
  filter(sector == 'Agriculture')
save(agriculture, file = "rda/agriculture.rdata")

### This part generates and saves a data set focused on crude oil

crude_oil <- energy %>%
  filter(commodity == 'Crude Oil')
save(crude_oil, file = "rda/crude_oil.rdata")

### This part generates and saves data sets focused on the commodities (other than crude oil) in the energy sector

natural_gas <- energy %>%
  filter(commodity == 'Natural Gas') %>%
  rename(natural_gas_closing_price = closing_price)
save(natural_gas, file = "rda/natural_gas.rdata")

heating_oil <- energy %>%
  filter(commodity == 'Heating Oil') %>%
  rename(heating_oil_closing_price = closing_price)
save(natural_gas, file = "rda/heating_oil.rdata")

gasoline <- energy %>%
  filter(commodity == 'Gasoline') %>%
  rename(gasoline_closing_price = closing_price)
save(gasoline, file = "rda/gasoline.rdata")

### This part generates and saves data sets focused on the commodities in the precious metals sector

gold <- precious_metals %>%
  filter(commodity == 'Gold') %>%
  rename(gold_closing_price = closing_price)
save(natural_gas, file = "rda/gold.rdata")

silver <- precious_metals %>%
  filter(commodity == 'Silver') %>%
  rename(silver_closing_price = closing_price)
save(natural_gas, file = "rda/silver.rdata")

platinum <- precious_metals %>%
  filter(commodity == 'Platinum') %>%
  rename(platinum_closing_price = closing_price)
save(gasoline, file = "rda/platinum.rdata")

### This part generates and saves data sets focused on the commodities in the agriculture sector

wheat <- agriculture %>%
  filter(commodity == 'Wheat') %>%
  rename(wheat_closing_price = closing_price)
save(wheat, file = "rda/wheat.rdata")

rice <- agriculture %>%
  filter(commodity == 'Rice') %>%
  rename(rice_closing_price = closing_price)
save(rice, file = "rda/rice.rdata")

soybeans <- agriculture %>%
  filter(commodity == 'Soybeans') %>%
  rename(soybeans_closing_price = closing_price)
save(soybeans, file = "rda/soybeans.rdata")

### This part generates a table that compares crude oil's closing price to the closing prices of all other commodities

natural_gas_price_table <- natural_gas %>% select(date, natural_gas_closing_price)
heating_oil_price_table <- heating_oil %>% select(date, heating_oil_closing_price)
gasoline_price_table <- gasoline %>% select(date, gasoline_closing_price)

gold_price_table <- gold %>% select(date, gold_closing_price)
silver_price_table <- silver %>% select(date, silver_closing_price)
platinum_price_table <- platinum %>% select(date, platinum_closing_price)

wheat_price_table <- wheat %>% select(date, wheat_closing_price)
rice_price_table <- rice %>% select(date, rice_closing_price)
soybeans_price_table <- soybeans %>% select(date, soybeans_closing_price)

crude_oil_in_commodities_market <- crude_oil %>%
  left_join(natural_gas_price_table) %>%
  left_join(heating_oil_price_table) %>%
  left_join(gasoline_price_table) %>%
  left_join(gold_price_table) %>%
  left_join(silver_price_table) %>%
  left_join(platinum_price_table) %>%
  left_join(wheat_price_table) %>%
  left_join(rice_price_table) %>%
  left_join(soybeans_price_table)

```

```{r conduct exploratory data analysis of crude oil in isolation, echo=FALSE}

### First, this script identitifies the maximum, minimum, mean, and standard deviation of crude oil closing prices for the 2010-2010 period of observation

crude_oil_price_at_beginning_of_period <- crude_oil$closing_price[1]
crude_oil_price_at_end_of_period <- crude_oil$closing_price[2519]
minimum_crude_oil_price_during_period <- min(crude_oil$closing_price)
maximum_crude_oil_price_during_period <- max(crude_oil$closing_price)
mean_crude_oil_price_during_period <- mean(crude_oil$closing_price)
standard_deviation_during_period <- sd(crude_oil$closing_price)

```

``` {r presentation of basic parameters describing crude oil closing prices, echo=TRUE}

print("Crude Oil Commodity Prices began the August 2010 to July 2020 period at")
crude_oil_price_at_beginning_of_period

print("They ended that ten-year period at")
crude_oil_price_at_end_of_period

print("During that time, they averaged")
mean_crude_oil_price_during_period

print("Running from a minimum of")
minimum_crude_oil_price_during_period

print("To a maximum of")
maximum_crude_oil_price_during_period

print("With a standard deviation of")
standard_deviation_during_period

```
``` {r plot of the 2010-2020 timeline of crude oil closing prices, echo=TRUE}

### Second, this script plots the 2010-2020 timeline of crude oil daily closing prices
### Then, it saves the plot in a .png file entitled "crude_oil_price_history"

crude_oil_plot <- crude_oil %>%
  ggplot(aes(date, closing_price)) +
  geom_point(aes(color = commodity)) +
  xlab("Date") +
  ylab("Closing Price") +
  ggtitle("Crude Oil Price History: 8/2010 - 7/2020") +
  theme_economist() +
    theme(legend.position="top",
          legend.title = element_blank(),
          legend.box = "horizontal" ,
          legend.text=element_text(size=8.5)) +
    guides(col = guide_legend(nrow = 1))

crude_oil_plot

ggsave("fig/crude_oil_price_history.png")

``` 

``` {r plots of crude oil closing prices by cyclical components of time, echo=TRUE}

### Third, this script explores crude oil closing prices in terms of the cyclical components of time
#### It plots the average crude oil closing prices by day of the trading week for 2010-2020
#### Then, it saves the plot in a .png file entitled "crude_oil_average_by_day_of_the_week"

crude_oil_average_by_day_of_the_week_plot <- crude_oil_in_commodities_market %>%
  group_by(date_weekday) %>%
  ggplot(aes(as.factor(date_weekday), closing_price)) +
  geom_boxplot(aes(color = commodity), size = 0.5, show.legend = FALSE) +
  xlab("Day of the Trading Week") +
  ylab("Closing Price") +
  ggtitle("Average Crude Oil Closing Price by Day of the Trading Week: 2010 - 2020") +
  theme_economist()

crude_oil_average_by_day_of_the_week_plot

ggsave("fig/crude_oil_average_by_day_of_the_week.png")

#### It plots the average crude oil closing prices by month of the year for 2010-2020
#### Then, it saves the plot in a .png file entitled "crude_oil_average_by_month_of_the_year

crude_oil_average_by_month_of_the_year_plot <- crude_oil_in_commodities_market %>%
  group_by(date_month_of_the_year) %>%
  ggplot(aes(as.factor(date_month_of_the_year), closing_price)) +
  geom_boxplot(aes(color = commodity), size = 0.5, show.legend = FALSE) +
  xlab("Month of the Year") +
  ylab("Closing Price") +
  ggtitle("Average Crude Oil Closing Price by Month of the Year: 2010 - 2020") +
  theme_economist()

crude_oil_average_by_month_of_the_year_plot

ggsave("fig/crude_oil_average_by_month_of_the_year.png")

#### It plots the average crude oil closing prices by quarter of the year for 2010-2020
#### Then, it saves the plot in a .png file entitled "crude_oil_average_by_month_of_the_year

crude_oil_average_by_quarter_of_the_year_plot <- crude_oil_in_commodities_market %>%
  group_by(date_quarter_of_the_year) %>%
  ggplot(aes(as.factor(date_quarter_of_the_year), closing_price)) +
  geom_boxplot(aes(color = commodity), size = 0.5, show.legend = FALSE) +
  xlab("Quarter of the Year") +
  ylab("Closing Price") +
  ggtitle("Average Crude Oil Closing Price by Quarter of the Year: 2010 - 2020") +
  theme_economist()

crude_oil_average_by_quarter_of_the_year_plot

ggsave("fig/crude_oil_average_by_quarter_of_the_year.png")

```

``` {r plots of crude oil closing prices by linear components of time, echo=TRUE}

### Fourth, this script explores crude oil closing prices in terms of the linear components of time
#### It plots the average crude oil closing prices by month for each month in 2010-2020
#### Then, it saves the plot in a .png file entitled "crude_oil_average_by_month"

crude_oil_average_by_month <- crude_oil_in_commodities_market %>%
  group_by(date_month) %>%
  mutate(month_avg = mean(closing_price))

crude_oil_average_by_month_plot <- crude_oil_average_by_month %>%
  ggplot(aes(date_month, month_avg)) +
  geom_point(aes(color = commodity), shape = 18, size = 3, show.legend = FALSE) +
  xlab("Month") +
  ylab("Closing Price") +
  ggtitle("Average Crude Oil Closing Price by Month: 2010 - 2020") +
  theme_economist()

crude_oil_average_by_month_plot

ggsave("fig/crude_oil_average_by_month.png")

#### It plots the average crude oil closing prices by year for each year in 2010-2020
#### Then, it saves the plot in a .png file entitled "crude_oil_average_by_year"

crude_oil_average_by_year <- crude_oil_in_commodities_market %>%
  group_by(date_year) %>%
  mutate(year_avg = mean(closing_price))

crude_oil_average_by_year_plot <- crude_oil_average_by_year %>%
  ggplot(aes(as.factor(date_year), year_avg)) +
  geom_point(aes(color = commodity), shape = 18, size = 5, show.legend = FALSE) +
  xlab("Year") +
  ylab("Closing Price") +
  ggtitle("Average Crude Oil Closing Price by Year: 2010 - 2020") +
  theme_economist()

crude_oil_average_by_year_plot

ggsave("fig/crude_oil_average_by_year.png")

```

``` {r plots of crude oil closing prices in comparison to other commodities, echo=TRUE}

### Fourth, this script explores crude oil closing prices in comparison to those of other complementary and competing commodities
#### It plots the 2010-2020 timeline of crude oil daily closing prices compared to those of the other energy sector commodities
#### Then, it saves the plot in a .png file entitled "crude_oil_price_history_in_comparison_to_three_other_energy_sector_commodities"

crude_oil_vs_other_energy_plot <- energy %>%
  ggplot(aes(date, closing_price)) +
  geom_point(aes(color = commodity)) +
  xlab("Date") +
  ylab("Closing Price") +
  ggtitle("Crude Oil versus Other Energy Sector Commodities") +
  theme_economist() +
  theme(legend.position="top",
        legend.title = element_blank(),
        legend.box = "horizontal" ,
        legend.text=element_text(size=8.5)) +
  guides(col = guide_legend(nrow = 1))

crude_oil_vs_other_energy_plot

ggsave("fig/crude_oil_in_comparison_to_three_other_energy_sector_commodities.png")

#### As a next step, it generates scatterplots portraying the relationships between crude oil closing prices and those of the other energy commodities

crude_oil_natural_gas_scatterplot <- crude_oil_in_commodities_market %>%
  ggplot(aes(closing_price, natural_gas_closing_price)) +
  geom_point(aes(color = commodity), size = 1, show.legend = FALSE) +
  xlab("Crude Oil Closing Price") +
  ylab("Natural Gas Closing Price") +
  ggtitle("Scatterplot of Crude Oil and Natural Gas Closing Prices: 2010 - 2020") +
  theme_economist() +
  theme(legend.position="top",
        legend.title = element_blank(),
        legend.box = "horizontal" ,
        legend.text=element_text(size=8.5)) +
  guides(col = guide_legend(nrow = 1))

crude_oil_natural_gas_scatterplot

ggsave("fig/scatterplot_crude_oil_versus_natural_gas.png")

crude_oil_heating_oil_scatterplot <- crude_oil_in_commodities_market %>%
  ggplot(aes(closing_price, heating_oil_closing_price)) +
  geom_point(aes(color = commodity), size = 1, show.legend = FALSE) +
  xlab("Crude Oil Closing Price") +
  ylab("Heating Oil Closing Price") +
  ggtitle("Scatterplot of Crude Oil and Heating Oil Closing Prices: 2010 - 2020") +
  theme_economist() +
  theme(legend.position="top",
        legend.title = element_blank(),
        legend.box = "horizontal" ,
        legend.text=element_text(size=8.5)) +
  guides(col = guide_legend(nrow = 1))

crude_oil_heating_oil_scatterplot

ggsave("fig/scatterplot_crude_oil_versus_heating_oil.png")

crude_oil_gasoline_scatterplot <- crude_oil_in_commodities_market %>%
  ggplot(aes(closing_price, gasoline_closing_price)) +
  geom_point(aes(color = commodity), size = 1, show.legend = FALSE) +
  xlab("Crude Oil Closing Price") +
  ylab("Gasoline Closing Price") +
  ggtitle("Scatterplot of Crude Oil and Gasoline Closing Prices: 2010 - 2020") +
  theme_economist() +
  theme(legend.position="top",
        legend.title = element_blank(),
        legend.box = "horizontal" ,
        legend.text=element_text(size=8.5)) +
  guides(col = guide_legend(nrow = 1))

crude_oil_gasoline_scatterplot

ggsave("fig/scatterplot_crude_oil_versus_gasoline.png")

### It plots the 2010-2020 timeline of crude oil daily closing prices compared to those of the three precious metals sector commodities
### Then, it saves the plot in a .png file entitled "crude_oil_price_history_in_comparison_to_three_competing_precious_metal_sector_commodities"

crude_oil_vs_precious_metals <- bind_rows(crude_oil, precious_metals)

crude_oil_vs_precious_metals_plot <- crude_oil_vs_precious_metals %>%
  ggplot(aes(date, closing_price)) +
  geom_point(aes(color = commodity)) +
  xlab("Date") +
  ylab("Closing Price") +
  ggtitle("Crude Oil versus Precious Metals") +
  theme_economist() +
  theme(legend.position="top",
        legend.title = element_blank(),
        legend.box = "horizontal" ,
        legend.text=element_text(size=8.5)) +
  guides(col = guide_legend(nrow = 1))

crude_oil_vs_precious_metals_plot

ggsave("fig/crude_oil_in_comparison_to_three_precious_metals_sector_commodities.png")

#### As a next step, it generates scatterplots portraying the relationships between crude oil closing prices and those of the precious metal commodities

crude_oil_gold_scatterplot <- crude_oil_in_commodities_market %>%
  ggplot(aes(closing_price, gold_closing_price)) +
  geom_point(aes(color = commodity), size = 1, show.legend = FALSE) +
  xlab("Crude Oil Closing Price") +
  ylab("Gold Closing Price") +
  ggtitle("Scatterplot of Crude Oil and Gold Closing Prices: 2010 - 2020") +
  theme_economist() +
  theme(legend.position="top",
        legend.title = element_blank(),
        legend.box = "horizontal" ,
        legend.text=element_text(size=8.5)) +
  guides(col = guide_legend(nrow = 1))

crude_oil_gold_scatterplot

ggsave("fig/scatterplot_crude_oil_versus_gold.png")

crude_oil_silver_scatterplot <- crude_oil_in_commodities_market %>%
  ggplot(aes(closing_price, silver_closing_price)) +
  geom_point(aes(color = commodity), size = 1, show.legend = FALSE) +
  xlab("Crude Oil Closing Price") +
  ylab("Silver Closing Price") +
  ggtitle("Scatterplot of Crude Oil and Silver Closing Prices: 2010 - 2020") +
  theme_economist() +
  theme(legend.position="top",
        legend.title = element_blank(),
        legend.box = "horizontal" ,
        legend.text=element_text(size=8.5)) +
  guides(col = guide_legend(nrow = 1))

crude_oil_silver_scatterplot

ggsave("fig/scatterplot_crude_oil_versus_silver.png")

crude_oil_platinum_scatterplot <- crude_oil_in_commodities_market %>%
  ggplot(aes(closing_price, gasoline_closing_price)) +
  geom_point(aes(color = commodity), size = 1, show.legend = FALSE) +
  xlab("Crude Oil Closing Price") +
  ylab("Platinum Closing Price") +
  ggtitle("Scatterplot of Crude Oil and Platinum Closing Prices: 2010 - 2020") +
  theme_economist() +
  theme(legend.position="top",
        legend.title = element_blank(),
        legend.box = "horizontal" ,
        legend.text=element_text(size=8.5)) +
  guides(col = guide_legend(nrow = 1))

crude_oil_platinum_scatterplot

ggsave("fig/scatterplot_crude_oil_versus_platinum.png")

### It plots the 2010-2020 timeline of crude oil daily closing prices compared to those of the three agricultural commodities
### Then, it saves the plot in a .png file entitled "crude_oil_price_history_in_comparison_to_three_competing_agriculture_sector_commodities"

crude_oil_vs_agriculture <- bind_rows(crude_oil, agriculture)

crude_oil_vs_agriculture_plot <- crude_oil_vs_agriculture %>%
  ggplot(aes(date, closing_price)) +
  geom_point(aes(color = commodity)) +
  xlab("Date") +
  ylab("Closing Price") +
  ggtitle("Crude Oil versus Agricultural Commodities") +
  theme_economist() +
  theme(legend.position="top",
        legend.title = element_blank(),
        legend.box = "horizontal" ,
        legend.text=element_text(size=8.5)) +
  guides(col = guide_legend(nrow = 1))

crude_oil_vs_agriculture_plot

ggsave("fig/crude_oil_in_comparison_to_three_agricultural_sector_commodities.png")

#### As a final step, it generates scatterplots portraying the relationships between crude oil closing prices and those of the agriculture commodities

crude_oil_wheat_scatterplot <- crude_oil_in_commodities_market %>%
  ggplot(aes(closing_price, gold_closing_price)) +
  geom_point(aes(color = commodity), size = 1, show.legend = FALSE) +
  xlab("Crude Oil Closing Price") +
  ylab("Wheat Closing Price") +
  ggtitle("Scatterplot of Crude Oil and Wheat Prices: 2010 - 2020") +
  theme_economist() +
  theme(legend.position="top",
        legend.title = element_blank(),
        legend.box = "horizontal" ,
        legend.text=element_text(size=8.5)) +
  guides(col = guide_legend(nrow = 1))

crude_oil_wheat_scatterplot

ggsave("fig/scatterplot_crude_oil_versus_wheat.png")

crude_oil_rice_scatterplot <- crude_oil_in_commodities_market %>%
  ggplot(aes(closing_price, rice_closing_price)) +
  geom_point(aes(color = commodity), size = 1, show.legend = FALSE) +
  xlab("Crude Oil Closing Price") +
  ylab("Rice Closing Price") +
  ggtitle("Scatterplot of Crude Oil and Rice Closing Prices: 2010 - 2020") +
  theme_economist() +
  theme(legend.position="top",
        legend.title = element_blank(),
        legend.box = "horizontal" ,
        legend.text=element_text(size=8.5)) +
  guides(col = guide_legend(nrow = 1))

crude_oil_rice_scatterplot

ggsave("fig/scatterplot_crude_oil_versus_rice.png")

crude_oil_soybeans_scatterplot <- crude_oil_in_commodities_market %>%
  ggplot(aes(closing_price, gasoline_closing_price)) +
  geom_point(aes(color = commodity), size = 1, show.legend = FALSE) +
  xlab("Crude Oil Closing Price") +
  ylab("Soybean Closing Price") +
  ggtitle("Scatterplot of Crude Oil and Soybean Closing Prices: 2010 - 2020") +
  theme_economist() +
  theme(legend.position="top",
        legend.title = element_blank(),
        legend.box = "horizontal" ,
        legend.text=element_text(size=8.5)) +
  guides(col = guide_legend(nrow = 1))

crude_oil_soybeans_scatterplot

ggsave("fig/scatterplot_crude_oil_versus_soybean.png")

```

```{r split the data into training and test sets, echo=FALSE}

### This part of the splits "crude_oil_in_commodities_market" into training and testing data sets

set.seed(755)

test_index <- createDataPartition(y = crude_oil_in_commodities_market$commodity, times = 1, p = 0.2, list = FALSE)

crude_oil_train <- crude_oil_in_commodities_market[-test_index,]

crude_oil_test <- crude_oil_in_commodities_market[test_index,]

```

```{r evaluation of algorithms, echo=TRUE}

### This section evaluates a set of fifteen potentially viable algorithms for predicting cruide oil closing prices

### Each of the first nine algorithms considers crude oil in isolation from other globally traded commodities
### In isolation, the available predictors include both the cyclical and the linear components of time
### One cyclical component is each observation's day of the trading week (Monday, Tuesday,...Friday)
### Others are the month (1,2..12) and the quarter (1,2,..4) of each year on the 2010-2020 timeline
### Liner components are the year (e.g., 2010) and month (e.g., 8/2010) of each observation

#### ALGORITHM 01 (AVERAGE)

##### This algorithm predicts crude oil closing prices based solely on the average crude oil price for 2010-2020
##### It will serve as a baseline from which to compare the analytical viability of other, more sophisticated algorithms

mu <- mean(crude_oil_train$closing_price)

predicted_price_algorithm_01 <- mu

RMSE01 <- RMSE(predicted_price_algorithm_01, crude_oil_test$closing_price)

RMSE01

```
